---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Load the data, unzipping it in-place. Convert date column into proper 
POSIXct type, parse the interval column as time, and combine date and time 
into timestamp.
```{r load_data, echo=TRUE}
library(dplyr)
library(lubridate)

data <- read.table(unz("activity.zip", "activity.csv"), 
                   header=TRUE, 
                   na.strings="NA", 
                   sep=",", 
                   as.is = TRUE)

# convert interval column to time and combine it with date column into timestamp
data <- data %>% mutate(date=ymd(date),
                        time=sprintf("%04d", interval), # add zero padding up to a width of 4
                        time=sub("(\\d{2})(\\d{2})", "\\1:\\2", time), # insert ":" between hours and minutes
                        timestamp=date + hm(time)) # convert to timestamp
str(data)
```



## What is mean total number of steps taken per day?
Let's group the data by date, calculate the sum of steps in each interval per group 
and plot a histogram.
```{r mean_daily_steps, echo=TRUE} 
library(ggplot2)
daily_data <- data %>% group_by(date) %>% summarise(steps_per_day=sum(steps, na.rm=TRUE))
ggplot(data=daily_data, aes(steps_per_day)) + geom_histogram()
mean_total_steps <- mean(daily_data$steps_per_day, na.rm=TRUE)
median_total_steps <- median(daily_data$steps_per_day, na.rm=TRUE)
```
The *mean* of the total steps per day is **`r mean_total_steps`**.
The *median* of the total steps per day is **`r median_total_steps`**.


## What is the average daily activity pattern?

To get a correct time-series plot we need to convert the interval column to proper
datetimes. Otherwise there would be "jumps" at the end of the hours; i.e. from 
1055 to 1100. Numerically the gap is 45, but time-wise it's only 5 minutes.

We do this by assigning the time to a single date (2015-09-20) and omitting
the date part in the axis labels.
```{r average_daily_pattern, echo=TRUE}
library(scales) # for date_format function
interval_data <- data %>% group_by(time) %>%
                      summarise(avg_steps=mean(steps, na.rm=TRUE)) %>%
                      mutate(time=ymd("2015-09-20") + hm(time))
ggplot(interval_data, aes(time, avg_steps)) + geom_line() + scale_x_datetime(labels = date_format("%H:%M"))
```

Let's find the index which contains the maximum average number of steps in an interval and obtain the
interval's corresponding start and end time.
```{r interval_max_steps, echo=TRUE}
max_ind <- which.max(interval_data$avg_steps)
start_time <- strftime(interval_data[[max_ind, "time"]], "%H:%M", tz="UTC") 
start_time
end_time <- strftime(interval_data[[max_ind+1, "time"]], "%H:%M", tz="UTC") 
end_time
```

The maximum number of steps accros all days occurs in the interval from **`r start_time`–`r end_time`**.

## Imputing missing values
```{r NAs, echo=TRUE}
num_NAs <- sum(is.na(data))
num_NAs
```
There are **`r num_NAs` missing values** in the dataset.

To fill in these missing values we use the mean for the corresponding time period.

```{r imputing, echo=TRUE}
avg_by_interval <- data %>% group_by(interval) %>% summarise(avg_steps=mean(steps, na.rm=TRUE))

# create a list with the same length as data and substitute with either the average of 
# the corresponding time period (if number of steps is NA) or with the given
# number of steps if the value is not NA. 
# The trick is that the avg_by_interval vector is reused (i.e. concatenated to itself) 
# if it is too short. Since every day in data has exactly 288 5-minute intervals
# (and avg_by_interval has 288 elements), this works as expected.
imp_steps <- ifelse(is.na(data$steps), avg_by_interval$avg_steps, data$steps)

# show that there are no NAs anymore
sum(is.na(data$imp_steps))

# column bind both
data <- cbind(data, imp_steps)
```

Let's group the data by date, calculate the sum of steps in each interval per group 
and plot a histogram.
```{r mean_daily_steps_v2, echo=TRUE} 
daily_data_v2 <- data %>% group_by(date) %>% summarise(steps_per_day=sum(imp_steps, na.rm=TRUE))
ggplot(data=daily_data_v2, aes(steps_per_day)) + geom_histogram()
mean_total_steps_v2 <- mean(daily_data_v2$steps_per_day, na.rm=TRUE)
median_total_steps_v2 <- median(daily_data_v2$steps_per_day, na.rm=TRUE)
```

The *mean* of the total steps per day is **`r prettyNum(mean_total_steps_v2)`**.
The *median* of the total steps per day is **`r prettyNum(median_total_steps_v2)`**.

Mean and median values are now identical. Mean value is higher than before, because
NAs got substiuted with the average of the corresponding interval over all days.

## Are there differences in activity patterns between weekdays and weekends?

```{r weekends, echo=TRUE, fig.width=10, fig.height=8}
# make sure the day of week labels appear in English
Sys.setlocale(category="LC_TIME", locale="en_US.UTF-8")

# create a factor variable for weekend/weekday
data <- data %>% mutate(weekend=factor(weekdays(data$timestamp, abbreviate = TRUE) %in% c("Sat","Sun"), levels=c(FALSE, TRUE), labels=c("weekday", "weekend")))

weekend_data <- data %>% group_by(weekend, time) %>%
                      summarise(avg_steps=mean(imp_steps, na.rm=TRUE)) %>%
                      mutate(time=ymd("2015-09-20") + hm(time))
ggplot(weekend_data, aes(time, avg_steps)) + 
    geom_line() + 
    facet_wrap(~weekend, ncol=1) +
    scale_x_datetime(labels = date_format("%H:%M"))

```

As can be seen, on average activity on weekends starts later and ends later. The
peek at 8:30 on weekdays seems to be the daily commute to work.
